REDO Simulating Dwarf Galaxy Evolution with Individual Star Feedback and Chemical Evolution I: Methods and Test Case

### 2017-11-20 main.tex
revisions through 2.5.1

We set the star formation density threshold to the point which the Jeans length becomes resolved by only 4 cells in sub-200 K gas, or about 200 \ccunit (as discussed further in Section~\ref{sec:star formation}). Forming stars from this gas will reduce the local density, ensuring the Jeans length is resolved. However, since star formation is not instantaneous, we employ a pressure floor to support gas against artificial collapse once its Jeans length becomes unresolved. This prevents dense, self-gravitating gas from rapidly reaching densities significantly above our resolution limit. The use of a pressure floor is common in galaxy scale simulations with limited dynamic range. (\textbf{I contrasted this to Hu's work which allows arbitrarily high densities, hoping feedback mitigates how high the densities actually get, but took it out. Leave it in?})

@@

We use the chemistry and cooling library \texttt{Grackle}\footnote{https://grackle.readthedocs.io/en/grackle-3.0/} v 3.0 to follow a nine species non-equilibrium chemistry network (H, H$^+$, He, He$^+$, He$^{++}$, e$^{-}$, H$_2$, H$^{-}$, and H$_{2}$) which includes radiative heating and cooling from these species and metals.\footnote{We use a slightly modified version of the the main \texttt{Grackle} repository, available at https://bitbucket.org/aemerick/grackle-emerick at changeset 59adc76b3130.} \texttt{Grackle} is a freely available, open source multi-code library, designed to interface with a wide variety of astrophysical codes. We outline specific model choices made in our simulations and refer the reader to \citep{GrackleMethod} for a detailed discussion of the code. We apply the \citet{Glover2008} three body rate for H$_{2}$ formation and include a model for H$_2$ formation on dust, dust heating, and dust cooling following the methods in \citet{2000ApJ...534..809O} and \citet{2005ApJ...626..627O} as included in \texttt{Grackle}. However, we update the default dust to gas ratio scaling in \texttt{Grackle} to account for the steeper scaling in low metallicity regimes ($Z \lesssim 0.1 Z_{\odot}$) with the broken power law scalings from \citet{Remy-Ruyer2014}. For metallicities above $\sim 0.1 Z_{\odot}$, this is equivalent to the the default behavior of \texttt{Grackle}, where dust content scales linearly with metallicity.
% Since we assume a dust abundance that depends only on metallicity, we do not track dust destruction or creation, which may play an important role regulating detailed ISM properties. (Vogelsberger is working on this directly but no papers yet... will cite if it gets submitted by then)

@@

\subsection{Star Formation Algorithm}
\label{sec:star formation}
In order resolve individual star formation on galaxy scales, we implement a stochastic star formation algorithm adopted from \citet{Goldbaum2015,Goldbaum2016}. Each cell at the maximum refinement level is capable of forming stars if it meets the following local criteria: 1) $n_{\rm cell} > n_{\rm thresh}$, 2) $T_{\rm gas} < T_{\rm thresh}$, 3) $M_{\rm gas} > M_{\rm Jeans}$, and 4) $\vec{\nabla} \cdot \vec{v} < 0$, where $n_{\rm thresh}$ is a resolution dependent density threshold, $T_{\rm thresh}$ is a temperature threshold, and $M_{\rm Jeans}$ is the local Jeans mass. Our fiducial values for $n_{\rm thresh}$ and $T_{\rm thresh}$ are 200~cm$^{-3}$ and 200~K respectively. We limit the fraction of a cell's gas mass that is converted into stars by requiring $M_{\rm gas} > f_{\rm thresh} M_{\rm max,*}$, where $f_{\rm thresh} = 2.0 $ and $M_{\rm max,*}$ is the maximum star mass (100 M$_{\odot}$). No star formation occurs when $M_{\rm gas} < f_{\rm thresh} M_{\rm max,*}$, ensuring that a star formation episode does not produce negative densities.

We make the common ansantz that star formation occurs by converting some fraction of gas into stars in a free fall time ($\tau_{\rm ff}$) as characterized by a star formation efficiency, $\epsilon_{\rm f} \simeq 0.02$. At high resolution, one's choice of $\epsilon_{\rm f}$ should be irrelevant \citep{Orr2017, FIRE2}, as star formation is ultimately self-regulated by feedback.
%but we find that $\epsilon_{\rm f}$ may impact star formation in the lowest mass halos, which are unusually sensitive to stellar ionization feedback (see Emerick et. al. in prep).
The the gas mass converted to stars in a time step, $dt$, is
\begin{equation}
M_{ {\rm gas} \to *} = \epsilon_{\rm f} M_{\rm gas} \frac{dt}{\tau_{\rm ff}}.
\end{equation}
In practice, $dt/\tau_{\rm ff} \ll 1$, and $M_{ {\rm gas} \to *}$ is smaller than the minimum star particle mass at parsec scale resolution. We therefore allow star formation to proceed stochastically, following the methods in \citet{Goldbaum2015, Goldbaum2016}, modified for variable stellar masses. In each cell that could potentially form stars, we compute the probability that 100 M$_{\odot}$ of gas will be converted into stars in that time step, and use a random number draw to determine whether or not star formation actually occurs. If it does, we random sample over the adopted IMF until 100 M$_{\odot}$ of stars form, keeping last sampled particle if the total stellar mass formed exceeds 100 M$_{\odot}$. We make the simple assumption of a Salpeter IMF \citep{Salpeter1955} with $\alpha = 2.35$, sampling over a minimum stellar mass of 1 M$_{\odot}$ and maximum stellar mass of 100 M$_{\odot}$.

Formed stars are deposited with random positions within the star forming cell and assigned velocities equal to the cell bulk velocity with a 1~km~s$^{-1}$ velocity dispersion. Stars are assigned metallicities corresponding to the metallicity of the star forming zone, and are chemically tagged with 17 individual species abundances (H, He, and the 15 metals) we follow in our simulations. Stars evolve throughout the simulation, losing mass from stellar winds and supernovae as described below, and by changing types, but persist throughout the entire simulation. For example, low mass stars are tagged as white dwarfs at the end of their life, which may eventually explode as a Type Ia supernova (discussed below), after which they persist as massless tracer particle ``remnants''.

@@

\subsection{Stellar Properties}
\label{sec:properties}
Given their birth mass and metallicity, we interpolate over the PARSEC grid of stellar evolution tracks \citep{Bressan2012} to assign a lifetime and AGB phase start time (if any) to each star, as well as an effective temperature ($T_{\rm eff}$) and surface gravity ($g$), used in computing radiation properties (see Section \ref{sec:ionizing radiation}). We use the largest subset of the PARSEC models that are regularly sampled in our mass/metallicity space of interest, with 26 mass bins over M$_{*} \in \left[0.95, 120.0 \right] {\rm M_{\odot}}$ and 11 metallicity bins over $Z \in \left[10^{-4}, 0.017 \right]$. Although $T_{\rm eff}$ and $g$ evolve over time for stars, modifying stellar radiative properties, following a stellar evolution track for each of our stars is beyond the scope of this work. We instead fix these properties at their zero age main sequence values.

@@

\subsubsection{Stellar Yields}
\label{sec:yields}
For the first time in galaxy scale simulations, we track galactic chemodynamical evolution using stellar yields as ejected from star particles that represent individual stars. We adopt the NuGrid\footnote{http://www.nugridstars.org} collaboration's set of stellar yields given on a uniformly sampled grid in stellar mass and metallicity with 12 mass bins over $M_{*} \in \left[1.0, 25.0\right]$ M$_{\odot}$ and five metallicity bins at $Z =$ 0.02, 0.01, 0.006, 0.001, and 10$^{-4}$ \citep{Pignatari2016, Ritter2017}. This grid includes yields from the AGB phase of stars between 1 and 7 M$_{\odot}$, as well as yields from both stellar winds and core collapse supernova of massive stars over 12 to 25 M$_{\odot}$. We compliment these tables to track stellar winds for stars more massive than 25 M$_{\odot}$ using tables from Slemer et. al. (in prep), based on the PARSEC stellar evolution tracks \citep{Bressan2012, Tang2014}; we ignore supernovae yields from these stars (see below). We combine all stable isotope yields for a given element into a single elemental abundance for all stable elements from hydrogen to bismuth. Although we can follow an arbitrary number of metal species, practical considerations prevent this in any given simulation. We refer the reader to previous uses of the NuGrid yields in one zone galactic chemical enrichment models \citep{Cote2016,  Cote2016_feb,Cote2017} for a detailed discussion on how various model uncertainties can influence galactic chemical evolution.

Above some mass, $M_{\rm trans}$, within the unsampled range of 7 to 12 M$_{\odot}$ stars do not undergo AGB wind phases and end their lives as core collapse supernovae. Where this transition occurs is uncertain, but is commonly taken to be around 8 - 10 M$_{\odot}$. This is a free parameter in our model, but we take $M_{\rm trans} = 8$ M$_{\odot}$ as our fiducial value. In our model, stars below this mass eject their wind yields in an AGB phase only at the end of their lives, typically over a period comparable to or less than a few time steps ($\lesssim 10$ kyr). Stars above this mass eject their stellar yields via ``massive" stellar winds at a constant mass loss rate throughout their lifetime, ending their lives as a core collapse supernova (see Section \ref{sec:stellar winds} for details on the wind energetics). Varying $M_{\rm thresh}$ changes both the time at which yields for stars around this mass are ejected (for reference, the lifetime of a 8 M$_{\odot}$ star is about 35 - 40 Myr), and the energy injection from these winds. \citet{Cote2017} explores how one's choice of $M_{\rm thresh}$ affects galaxy abundances in a one-zone model.

There are large uncertainties in stellar yields for stars more massive than 25 M$_{\odot}$ \citep[see ][and references therein]{Cote2016}. Indeed, even the exact fate of these stars is uncertain \citep[e.g.][]{Woosley2002,Zhang2008,Ugliano2012}, particularly as a function of metallicity \citep{Fryer2012} with potentially multiple stable/unstable regimes as a function of mass \citep{Heger2003}. Due to this uncertainty, and to avoid erroneously extrapolating from our yield tables, we adopt the simplest model and assume all stars above 25 M$_{\odot}$ end their life through direct collapse to a compact object with no mass or energy ejection.

Type Ia supernova are an important additional source of galactic chemical enrichment. These iron group rich events are responsible for the $\sim$1 Gyr timescale turnover, or ``knee", in $[\alpha/\rm{Fe}]$ vs $[\rm{Fe}/\rm{H}]$ diagrams. We use the Type Ia supernova yields given in \citet{Thielemann1986}, adopting a Type Ia supernova model as discussed in Section~\ref{sec:Type Ia}.

\subsubsection{Stellar Winds}
\label{sec:stellar winds}
Stellar winds are important sources of early (massive stars) and late (AGB stars) enrichment and feedback in galaxies. Although the energy injected via winds over the lifetime of a cluster of stars is much less than that of supernovae and radiation, stellar winds are potentially important sources of pre-supernova feedback. In both AGB phase winds and massive star winds we inject the feedback energy in a 3-cell radius spherical region centered on the emitting star particle assuming full thermalization of the wind kinetic energy. The total injected energy in timestep $\Delta t$ is therefore $E_{\rm wind} = \dot{M}\Delta t \times v^2_{\rm wind} + E_{\rm th}$, where $E_{\rm th}$ is the thermal energy of the ejected gas mass ($\dot{M}\Delta t$) given the star's interpolated $T_{\rm eff}$. We use a monte-carlo sampling method to inject mass and energy into the surrounding cells to account for partial cell overlap at the edge if the injection region that is a result of mapping a spherical injection region to cubic grid cells.

>>>>

We set the star formation density threshold to the value at which the Jeans length becomes resolved by only 4 cells in sub-200 K gas, or about 200 \ccunit (as discussed further in Section~\ref{sec:star formation}). Forming stars from this gas will reduce the local density, ensuring the Jeans length is resolved. However, since star formation is not instantaneous, we employ a pressure floor to support gas against artificial collapse once its Jeans length becomes unresolved. This prevents dense, self-gravitating gas from rapidly reaching densities significantly above our resolution limit. The use of a pressure floor is common in galaxy scale simulations with limited dynamic range. (\textbf{I contrasted this to Hu's work which allows arbitrarily high densities, hoping feedback mitigates how high the densities actually get, but took it out. Leave it in?})

@@

We use the chemistry and cooling library \texttt{Grackle}\footnote{https://grackle.readthedocs.io/en/grackle-3.0/} v 3.0 to follow a nine species non-equilibrium chemistry network (H, H$^+$, He, He$^+$, He$^{++}$, e$^{-}$, H$_2$, H$^{-}$, and H$_{2}$) which includes radiative heating and cooling from these species and metals.\footnote{We use a slightly modified version of the the main \texttt{Grackle} repository, available at https://bitbucket.org/aemerick/grackle-emerick at changeset 59adc76b3130.} \texttt{Grackle} is a freely available, open source multi-code library, designed to interface with a wide variety of astrophysical codes. We outline specific model choices made in our simulations and refer the reader to \citet{GrackleMethod} for a detailed discussion of the code. We apply the \citet{Glover2008} three-body rate for H$_{2}$ formation and include a model for H$_2$ formation on dust, dust heating, and dust cooling following the methods in \citet{2000ApJ...534..809O} and \citet{2005ApJ...626..627O} as included in \texttt{Grackle}. However, we update the default dust to gas ratio scaling in \texttt{Grackle} to account for the steeper scaling in low metallicity regimes ($Z \lesssim 0.1 Z_{\odot}$) with the broken power law scalings from \citet{Remy-Ruyer2014}. For metallicities above $\sim 0.1 Z_{\odot}$, this is equivalent to the the default behavior of \texttt{Grackle}, where dust content scales linearly with metallicity.
% Since we assume a dust abundance that depends only on metallicity, we do not track dust destruction or creation, which may play an important role regulating detailed ISM properties. (Vogelsberger is working on this directly but no papers yet... will cite if it gets submitted by then)

@@

\subsection{Star Formation Algorithm}
\label{sec:star formation}
In order to resolve individual star formation on galaxy scales, we implement a stochastic star formation algorithm adopted from \citet{Goldbaum2015,Goldbaum2016}. Each cell at the maximum refinement level is capable of forming stars if it meets the following local criteria: 1) $n_{\rm cell} > n_{\rm thresh}$, 2) $T_{\rm gas} < T_{\rm thresh}$, 3) $M_{\rm gas} > M_{\rm Jeans}$, and 4) $\vec{\nabla} \cdot \vec{v} < 0$, where $n_{\rm thresh}$ is a resolution dependent density threshold, $T_{\rm thresh}$ is a temperature threshold, and $M_{\rm Jeans}$ is the local Jeans mass. Our fiducial values for $n_{\rm thresh}$ and $T_{\rm thresh}$ are 200~cm$^{-3}$ and 200~K respectively. We limit the fraction of a cell's gas mass that is converted into stars by requiring $M_{\rm gas} > f_{\rm thresh} M_{\rm max,*}$, where $f_{\rm thresh} = 2.0 $ and $M_{\rm max,*}$ is the maximum star mass (100 M$_{\odot}$). No star formation occurs when $M_{\rm gas} < f_{\rm thresh} M_{\rm max,*}$, ensuring that a star formation episode does not produce negative densities.

We make the common ansatz that star formation occurs by converting some fraction of gas into stars in a free fall time ($\tau_{\rm ff}$) as characterized by a star formation efficiency, $\epsilon_{\rm f} \simeq 0.02$. At high resolution, one's choice of $\epsilon_{\rm f}$ should be irrelevant \citep{Orr2017, FIRE2}, as star formation is ultimately self-regulated by feedback.
%but we find that $\epsilon_{\rm f}$ may impact star formation in the lowest mass halos, which are unusually sensitive to stellar ionization feedback (see Emerick et. al. in prep).
The gas mass converted to stars in a time step, $dt$, is
\begin{equation}
M_{ {\rm gas} \to *} = \epsilon_{\rm f} M_{\rm gas} \frac{dt}{\tau_{\rm ff}}.
\end{equation}
In practice, $dt/\tau_{\rm ff} \ll 1$, and $M_{ {\rm gas} \to *}$ is smaller than the minimum star particle mass at parsec scale resolution. We therefore allow star formation to proceed stochastically, following the methods in \citet{Goldbaum2015, Goldbaum2016}, modified for variable stellar masses. In each cell that could potentially form stars, we compute the probability that 100 M$_{\odot}$ of gas will be converted into stars in that time step, and use a random number draw to determine whether or not star formation actually occurs. If it does, we random sample over the adopted IMF until 100 M$_{\odot}$ of stars form, keeping the last sampled particle if the total stellar mass formed exceeds 100 M$_{\odot}$. We make the simple assumption of a Salpeter IMF \citep{Salpeter1955} with $\alpha = 2.35$, sampling over a minimum stellar mass of 1 M$_{\odot}$ and maximum stellar mass of 100 M$_{\odot}$.

Formed stars are deposited with random positions within the star forming cell and assigned velocities equal to the cell bulk velocity with a 1~km~s$^{-1}$ velocity dispersion. Stars are assigned metallicities corresponding to the metallicity of the star forming zone, and are chemically tagged with the 17 individual species abundances (H, He, and the 15 metals) that we follow in our simulations. Stars evolve during the simulation, by losing mass from stellar winds and supernovae as described below, and by changing types, but persist throughout the entire simulation. For example, low mass stars are tagged as white dwarfs at the end of their life, which may eventually explode as a Type Ia supernova (discussed below), after which they persist as massless tracer particle ``remnants''.

@@

\subsection{Stellar Properties}
\label{sec:properties}
Given each star's birth mass and metallicity, we interpolate over the PARSEC grid of stellar evolution tracks \citep{Bressan2012} to assign a lifetime and AGB phase start time (if any) to it, as well as the effective temperature $T_{\rm eff}$ and surface gravity $g$ used in computing radiation properties (see Section \ref{sec:ionizing radiation}). We use the largest subset of the PARSEC models that are regularly sampled in our mass/metallicity space of interest, with 26 mass bins over M$_{*} \in \left[0.95, 120.0 \right] {\rm M_{\odot}}$ and 11 metallicity bins over $Z \in \left[10^{-4}, 0.017 \right]$. Although $T_{\rm eff}$ and $g$ evolve over time for stars, modifying stellar radiative properties, following a stellar evolution track for each of our stars is beyond the scope of this work. We instead fix these properties at their zero age main sequence values.

@@

\subsubsection{Stellar Yields}
\label{sec:yields}
For the first time in galaxy scale simulations, we track galactic chemodynamical evolution using stellar yields as ejected from star particles that represent individual stars. We adopt the NuGrid\footnote{http://www.nugridstars.org} collaboration's set of stellar yields given on a uniformly sampled grid in stellar mass and metallicity with 12 mass bins over $M_{*} \in \left[1.0, 25.0\right]$ M$_{\odot}$ and five metallicity bins at $Z =$ 0.02, 0.01, 0.006, 0.001, and 10$^{-4}$ \citep{Pignatari2016, Ritter2017}. This grid includes yields from the AGB phase of stars between 1 and 7 M$_{\odot}$, as well as yields from both stellar winds and core collapse supernova of massive stars over 12 to 25 M$_{\odot}$. We complement these tables with tables from Slemer et. al. (in prep), based on the PARSEC stellar evolution tracks \citep{Bressan2012, Tang2014} to track stellar winds for stars more massive than 25 M$_{\odot}$. We ignore supernova yields from these stars (see next paragraph). We combine all stable isotope yields for a given element into a single elemental abundance for all stable elements from hydrogen to bismuth. Although we can follow an arbitrary number of metal species, practical considerations 
%mm 
   of memory use 
prevent this in any given simulation. We refer the reader to previous uses of the NuGrid yields in one-zone galactic chemical enrichment models \citep{Cote2016,  Cote2016_feb,Cote2017} for a detailed discussion of how various model uncertainties can influence galactic chemical evolution.

Above some mass $M_{\rm trans}$ within the unsampled range of 7--12 M$_{\odot}$, stars do not undergo AGB wind phases and end their lives as core collapse supernovae. Where this transition occurs is uncertain, but is commonly taken to be around 8--10~M$_{\odot}$. This is a free parameter in our model, but we take $M_{\rm trans} = 8$~M$_{\odot}$ as our fiducial value. In our model, stars below this mass eject their wind yields in an AGB phase only at the end of their lives, typically over a period comparable to or less than a few time steps ($\lesssim 10$ kyr). Stars above this mass 
%mm
    are assumed to
eject their stellar yields via 
%mm ``massive" 
   line-driven
stellar winds at a constant mass loss rate throughout their lifetime
%mm ,
  (neglecting Wolf-Rayet and luminous blue variable phases),
ending their lives as a core-collapse supernova (see Section \ref{sec:stellar winds} for details on the wind energetics). Varying $M_{\rm thresh}$ changes both the time at which yields for stars around this mass are ejected (for reference, the lifetime of a 8 M$_{\odot}$ star is about 35--40~Myr), and the energy injection from these winds. \citet{Cote2017} explores how the choice of $M_{\rm thresh}$ affects galaxy abundances in a one-zone model.

There are large uncertainties in stellar yields for stars more massive than 25 M$_{\odot}$ \citep[see ][and references therein]{Cote2016}. Indeed, even the exact fate of these stars is uncertain \citep[e.g.][]{Woosley2002,Zhang2008,Ugliano2012}, particularly as a function of metallicity \citep{Fryer2012} with potentially multiple stable and unstable regimes as a function of mass \citep{Heger2003}. Due to this uncertainty, and to avoid erroneously extrapolating from our yield tables, we adopt the simplest model and assume all stars above 25 M$_{\odot}$ end their life through direct collapse to a compact object with no 
%mm
    further
mass or energy ejection.

Type Ia supernovae are an important additional source of galactic chemical enrichment. These iron group rich events are responsible for the $\sim$1 Gyr timescale turnover, or ``knee", in $[\alpha/\rm{Fe}]$ vs $[\rm{Fe}/\rm{H}]$ diagrams. We use the Type Ia supernova yields given in \citet{Thielemann1986}, adopting a Type Ia supernova model as discussed in Section~\ref{sec:Type Ia}. 
%mm
    We emphasize that we only track Type Ia supernovae occurring within the 
    population of stars formed in this model, neglecting supernovae from any 
    pre-existing population.

\subsubsection{Stellar Winds}
\label{sec:stellar winds}
Stellar winds are important sources of enrichment and feedback in galaxies at both early times from massive stars and late times from AGB stars. Although the energy injected via winds over the lifetime of a cluster of stars is much less than that of supernovae and radiation, stellar winds are potentially important sources of pre-supernova feedback. In both AGB phase winds and massive star winds we inject the feedback energy in a 3-cell radius spherical region centered on the emitting star particle assuming full thermalization of the wind kinetic energy. The total injected energy in timestep $\Delta t$ is therefore $E_{\rm wind} = \dot{M}\Delta t \times v^2_{\rm wind} + E_{\rm th}$, where $E_{\rm th}$ is the thermal energy of the ejected gas mass ($\dot{M}\Delta t$) given the star's interpolated $T_{\rm eff}$. We use a monte-carlo sampling method to inject mass and energy into the surrounding cells to account for partial cell overlap at the edge if the injection region that is a result of mapping a spherical injection region to cubic grid cells.
